---
title: "Data Analysis Project"
author: "Sandra Smith, Lauren Bentley, Katelynn Winebrenner"
date: "May 13, 2025"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction, Preliminary Analysis and Definitions

In this notebook, we are analyzing Maryland precinct data from the 2020 and 2024 elections. To do this, we have loaded in 2 data sets from the Maryland State Board of Elections. 

The 2020 data set has 23 columns and 95283 rows. 
The 2024 data set has 21 columns and 117875 rows. 
Both data sets show each precinct, information about the precinct, each race, each candidate, the winner, how many votes each candidate received and method of vote. 

Because each year has different candidates and different races, we should identify which races happened both years and in all counties in order to make a clear analysis. We should also identify clear candidates in each race so that we aren't analyzing someone who is listed, but received one write in vote, for example.

To meet these parameters, we will focus on the race for president because it happened both years and in every precinct. In terms of candidates, we will focus on Donald Trump and Joe Biden for 2020 and Kamala Harris and Donald Trump for 2024.

We should be able to answer most of our questions with this data. However, we will need to load in data from the census in order to answer questions about economic status and identify. I don't know how easy that will be to do by precinct, but we'll see. We'll also need to include information about eligible voters and registered voters to determine turnout.


## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}

options(scipen=999)
library(lubridate)
library(tidycensus)
library(janitor)
library(sf)
library(mapview)
library(tigris)
library(lubridate)
library(tidyverse)
```

## Load and Clean Data
```{r}
# Load required data
uncleaned_precincts2020 <- read_csv("data/All_By_Precinct_2020_General.csv")
uncleaned_precincts2024 <- read_csv("data/PG24_AllPrecincts.csv")
```

```{r}
# Clean 2020, column names to match, county names, combine column district and column precinct, select columns
precincts2020 <- uncleaned_precincts2020 |>
  clean_names() |>
  mutate(county_name = case_when(
    county == "01" ~ "Allegany County",
    county == "02" ~ "Anne Arundel County",
    county == "03" ~ "Baltimore City",
    county == "04" ~ "Baltimore County",
    county == "05" ~ "Calvert County",
    county == "06" ~ "Caroline County",
    county == "07" ~ "Carroll County",
    county == "08" ~ "Cecil County",
    county == "09" ~ "Charles County",
    county == "10" ~ "Dorchester County",
    county == "11" ~ "Frederick County",
    county == "12" ~ "Garrett County",
    county == "13" ~ "Harford County",
    county == "14" ~ "Howard County",
    county == "15" ~ "Kent County",
    county == "16" ~ "Montgomery County",
    county == "17" ~ "Prince George's County",
    county == "18" ~ "Queen Anne's County",
    county == "19" ~ "Saint Mary's County",
    county == "20" ~ "Somerset County",
    county == "21" ~ "Talbot County",
    county == "22" ~ "Washington County",
    county == "23" ~ "Wicomico County",
    county == "24" ~ "Worcester County"
  ))|>
  mutate(election_district_precinct = str_c(election_district, election_precinct, sep = "-"),
         year = 2020)|> #chatgpt, str_c function to combine two column outputs by - in single column
  rename(
    election_day = election_day_votes,
    early= early_voting_votes,
    mail_in_1= by_mail_votes,
    mail_in_2 = by_mail_2_votes,
    provisional= prov_votes,
    total= total_votes,
    candidate_name = candidate_name,
    party = party
         )|>
      select(year,county, county_name, election_district_precinct, office_name, candidate_name, party, election_day, early, mail_in_1, mail_in_2, provisional, total)|>
  filter(
    office_name == "President - Vice Pres", 
    candidate_name %in% c("Joe Biden", "Donald J. Trump")) #%n% from chatgpt


# Clean 2024, column names to match, create total votes, select columns
precincts2024 <- uncleaned_precincts2024|>
  clean_names()|>
   rename(
     election_day = election_night_votes,
     mail_in_1 = mail_in_ballot_1_votes,
     mail_in_2 = mail_in_ballot_2_votes,
     early = early_votes,
     provisional= provisional_votes,
     candidate_name = candidate_name,
     party = party
     )|>
  mutate(
    total = (election_day + mail_in_1 + mail_in_2 + early + provisional),
    year = 2024
  )|>
    select(year, county, county_name, election_district_precinct, office_name, candidate_name, party, election_day, early, mail_in_1, mail_in_2, provisional, total)|>
 filter(
    office_name == "President - Vice Pres",
    candidate_name %in% c("Kamala D. Harris and Tim Walz", "Donald J. Trump and JD Vance") #%n% from chatgpt
  )

```

```{r}
# combine dataframes, not joining, but placing them on top of one another
precincts_combined <- bind_rows(precincts2020, precincts2024)

```


## Questions to answer
In this notebook, we are answering five questions about the data.




### Question 2

* **Question**: How did voter turnout vary across precincts? What factors contributed to this? 

* **Analysis summary**: 

### Question 1

* **Question**: Which precincts experienced the most significant change in party support? 

* **Analysis summary**: 

```{r}
#Summarize votes by party (2020)
votes_2020 <- precincts_combined |>
  filter(year == 2020) |>
  group_by(election_district_precinct, county_name, party) |>
  summarize(total_votes_2020 = sum(total,na.rm = TRUE), .groups= "drop")

#Summarize votes by party (2024)
votes_2024 <- precincts_combined |>
  filter(year == 2024) |>
  group_by(election_district_precinct, county_name, party) |>
  summarize(total_votes_2024 = sum(total, na.rm=TRUE), .groups="drop")

#Join 2020 and 2024 datasets by party and precinct 
vote_changes <- left_join(votes_2020, votes_2024, by = c("election_district_precinct", "county_name", "party"))

#Democrat percent change 
democratic_changes <- vote_changes |>
  filter(party == "DEM") |>
  mutate(
    percent_change = ((total_votes_2024 - total_votes_2020) / total_votes_2020) * 100
  ) |>
  select(election_district_precinct, county_name, percent_change) |>
  arrange(desc(percent_change))

#Republican percent change 
republican_changes <- vote_changes |>
  filter(party == "REP") |>
  mutate(
    percent_change = ((total_votes_2024 - total_votes_2020) / total_votes_2020) * 100 
  ) |>
  select(election_district_precinct, county_name, percent_change) |>
  arrange(desc(percent_change))
```







### Question 3

* **Question**: How did early voting and mail-in ballots influence precinct-level results? 


* **Analysis summary**: 

```{r}

```



### Question 4

* **Question**: How did voting behavior differ between precincts based on economic status? 

* **Analysis summary**: 



```{r}
precincts_shape <- st_read("data/Maryland_Election_Boundaries_-_Precincts_2022/Maryland_Election_Boundaries_-_Precincts_2022.shp")|>
  clean_names()|>
   mutate(
    name = str_extract(name, "\\d{1,2}-\\d{3}"),  # Extract the district-precinct code (e.g., "29-002")
    district = str_extract(name, "^\\d{1,2}"),  # Extract the district number (e.g., "29", "5", "12")
    precinct_code = str_pad(district, width = 3, side = "left", pad = "0"),  # Pad district with leading zeros
    name = str_replace(name, "^\\d{1,2}", precinct_code)  # Replace original district with padded district
  ) |>
rename(election_district_precinct = name)


#join precinct data to geo data

geometry_precincts_2020 <- precincts_shape |>
  left_join(precincts2020, join_by(election_district_precinct)) 

geometry_precincts_2024 <- precincts_shape |>
  left_join(precincts2024, join_by(election_district_precinct)) 

#load census tracts, income data
median_income_pop <- get_acs(
  geography = "tract",
  variables = "B19013_001",
  state = "MD",
  year = 2021,
  geometry = TRUE,
  ) |> 
  clean_names()
  
  

median_income_pop |>
  as.data.frame()



```

```{r error=TRUE}
geometry_precincts_2020 <- geometry_precincts_2020 |>
  st_transform(crs=5070)

geometry_precincts_2024 <- geometry_precincts_2024 |>
  st_transform(crs=5070)

median_income_pop <- median_income_pop|>
  st_transform(crs=5070)


# Perform a spatial join using st_within
#spatial_join_2020 <- st_within(geometry_precincts_2020, median_income_pop, sparse = FALSE)

#spatial_join_2024  <- st_within(geometry_precincts_2024, median_income_pop, sparse = FALSE)

```

```{r}
MAP :()
map <- mapview(geometry_precincts_2020, 
               col.region = "blue", + # Set the color of the precincts
       mapview(median_income_pop, 
                     col.region = "red", # Set the color of the census tracts
                     alpha.region = 0.8,
                     cex = 2, # Set the transparency
                     legend = TRUE) # Enable the legend

view(map)
```



### Question 5

* **Question**: How did voting behavior differ between precincts based on race in 2020 and 2024?

* **Analysis summary**: 

In 2020, Trump won in about 532 Maryland precincts; 98.5% of those precincts have a white racial majority. He won in two majority Hispanic precincts.

In 2020, Biden won in about 1429 Maryland precincts; 37% of those precincts have a white racial majority, 33.8% have Black racial majority and 26% have no racial majority. He won in 37 majority Hispanic precincts.

In 2024, Trump won in about 481 Maryland precincts; 96% of those have a white racial majority. He won in one majority Hispanic precinct.

In 2024, Harris won in about 1318 Maryland precincts; 36% of those have a white racial majority, 35.8% have a black majority and 25% have no racial majority. She won in 35 majority Hispanic precincts.



```{r}
#assign racial majority to precincts based on Maryland Planning Board data:  https://planning.maryland.gov/Redistricting/Pages/2020/precinct.aspx

uncleaned_precinct_population <- read_csv("data/precinct_population_data.csv") 

precinct_population<- uncleaned_precinct_population|>
  clean_names() |>
  select(
    precinct_name, voting_district,
    white, black, native_american, asian, pacific_islander, other,
    two_or_more, hispanic_latino
  )|>
  rename(county_name = precinct_name)|>
  mutate(
    election_district_precinct = (paste0("0", voting_district)),
    county_name = str_remove(county_name, " Precinct.*"),  # Remove everything after ' Precinct'
    county_name = case_when(county_name == "Baltimore County " ~ "Baltimore",
                            TRUE ~ county_name),
    county_name = case_when(county_name == "St. Mary's" ~ "Saint Mary's",
                            TRUE ~ county_name),
    county_name = paste0(county_name, " County"),  # Add 'County' at the end
    county_name = case_when(county_name == "Baltimore City County" ~ "Baltimore City",
                            TRUE ~ county_name),
    total_population =  (white + black + native_american + asian + pacific_islander + other + 
    two_or_more + hispanic_latino),
    pct_white = (white/total_population)*100,
    pct_black =(black/total_population)*100,
    pct_asian = (asian/total_population)*100,
    pct_hispanic = (hispanic_latino/total_population )*100,
    race = case_when( 
      pct_white >= 50 ~ "Majority White",
      pct_black >= 50 ~ "Majority Black",
      pct_asian >= 50 ~ "Majority Asian",
      pct_hispanic >= 50 ~"Majority Hispanic",
       TRUE ~ "No Racial Majority"
      ))|>
  select(county_name, election_district_precinct,race)
  
trump2020<- precincts2020 |>
  filter(candidate_name == "Donald J. Trump")|>
  select(county_name, election_district_precinct,total)|>
  rename(trump_total_votes=total)
    

biden2020<- precincts2020 |>
  filter(candidate_name == "Joe Biden")|>
  select(county_name, election_district_precinct,total)|>
  rename(biden_total_votes=total)

election2020 <- trump2020|>
  left_join(biden2020, join_by(county_name, election_district_precinct))|>
  mutate(
    # Replace NA vote counts with 0 -- ChatGPT
    trump_total_votes = replace_na(trump_total_votes, 0),
    biden_total_votes = replace_na(biden_total_votes, 0),
    total_votes = biden_total_votes + trump_total_votes,
    # Prevent divide-by-zero by using if_else -- ChatGPT
    pct_trump = if_else(total_votes == 0, NA_real_, (trump_total_votes / total_votes) * 100),
    pct_biden = if_else(total_votes == 0, NA_real_, (biden_total_votes / total_votes) * 100),
    winner= case_when(
      pct_trump>=50 ~ "Trump",
      pct_biden>=50 ~"Biden",
      TRUE ~ "No Majority Winner"
    )
  )|>
  select(county_name,election_district_precinct,winner)
  
#add racial data to 2020 precinct election data
  election_population_2020 <- precinct_population |>
    left_join( election2020, join_by(county_name,election_district_precinct))|>
    group_by(winner,race)|>
    summarise(
      precincts = n()
    )|>
    filter(!is.na(winner))|>
    mutate(
      total_precincts_won = sum(precincts)
    )|>
    arrange(desc(total_precincts_won))
  
  racial_analysis2020 <- election_population_2020|>
    mutate(
      precinct_pct = (precincts/ total_precincts_won)*100
    )
#run analysis
  racial_analysis2020
    
# repeat for 2024
trump2024<- precincts2024 |>
  filter(candidate_name == "Donald J. Trump and JD Vance")|>
  select(county_name, election_district_precinct,total)|>
  rename(trump_total_votes=total)
    
harris2024 <- precincts2024 |>
  filter(candidate_name == "Kamala D. Harris and Tim Walz")|>
  select(county_name, election_district_precinct,total)|>
  rename(harris_total_votes=total)

election2024 <- trump2024|>
  left_join(harris2024, join_by(county_name, election_district_precinct))|>
  mutate(
    # Replace NA vote counts with 0 -- Chat GPT
    trump_total_votes = replace_na(trump_total_votes, 0),
    harris_total_votes = replace_na(harris_total_votes, 0),
    total_votes = harris_total_votes + trump_total_votes,
    # Prevent divide-by-zero by using if_else -- Chat GPT
    pct_trump = if_else(total_votes == 0, NA_real_, (trump_total_votes / total_votes) * 100),
    pct_harris = if_else(total_votes == 0, NA_real_, (harris_total_votes / total_votes) * 100),
    winner= case_when(
      pct_trump>=50 ~ "Trump",
      pct_harris>=50 ~"Harris",
      TRUE ~ "No Majority Winner"
    )
  )|>
  select(county_name,election_district_precinct,winner)
  
#add racial data to 2020 precinct election data
  election_population_2024 <- precinct_population |>
    left_join( election2024, join_by(county_name,election_district_precinct))|>
    group_by(winner,race)|>
    summarise(
      precincts = n()
    )|>
    filter(!is.na(winner))|>
    mutate(
      total_precincts_won = sum(precincts)
    )|>
    arrange(desc(total_precincts_won))
  
  racial_analysis2024 <- election_population_2024|>
    mutate(
      precinct_pct = (precincts/ total_precincts_won)*100
    )|>
    arrange(desc(precinct_pct))
  
#run analysis
    racial_analysis2024
 
```



