---
title: "lab_02"
author: "derek willis"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland.

**A1** There appears to be 17099 new registrants since July 21. Montgomery County saw the most in total.	



```{r}
 BIDEN_DROPOUT_DATE <- ymd(20240721)

### ONE_DAY_FOLLOWING <- ymd(20240722)
### BIDEN_DROPOUT_DATE - ONE_DAY_FOLLOWING = -1 days 

md_voters_2024 |>
  mutate( NEW_REGISTRATION = case_when(
   STATE_REGISTRATION_DATE - COUNTY_REGISTRATION_DATE == 0 
   & 
   BIDEN_DROPOUT_DATE - COUNTY_REGISTRATION_DATE <= 0 ~ "NEW Post-Biden",
  )
  )|>


group_by(COUNTY) |>
 filter(NEW_REGISTRATION == "NEW Post-Biden") |>
summarize(
 POST_BIDEN_REGISTRATIONS = n()
)|>
arrange(desc(POST_BIDEN_REGISTRATIONS))

```




**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2** The weeks from July 21-August 7 do see a greater number of new registrations in general, especially the week of July 21st. The only other weeks before this with high first-time registrations was the week of June 9 and July 7. All other weeks are below 5,000.

```{r}
md_voters_2024 |>
  
  mutate(week = floor_date(COUNTY_REGISTRATION_DATE, "week"),
        NEW_REGISTRATION = case_when(
          STATE_REGISTRATION_DATE - COUNTY_REGISTRATION_DATE == 0 
          ~ "NEW_VOTER")
         
         ) |>
  
  group_by(week) |>
 filter(NEW_REGISTRATION == "NEW_VOTER") |>
  summarise(
    TOTAL_NEW_VOTERS = n()) |>
  arrange(desc(TOTAL_NEW_VOTERS))




```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**A3** July was the month with the greatest number of new registrations , this sheds more light on the previous question because there is a very big jump between June, which is second, and July.

```{r}
md_voters_2024 |>
  
  mutate(month = floor_date(COUNTY_REGISTRATION_DATE, "month"),
        NEW_REGISTRATION = case_when(
          STATE_REGISTRATION_DATE - COUNTY_REGISTRATION_DATE == 0 
          ~ "NEW_VOTER")
         
         ) |>
  
  group_by(month) |>
 filter(NEW_REGISTRATION == "NEW_VOTER")|>
  summarise(
    TOTAL_NEW_VOTERS = n()) |>
  arrange(desc(TOTAL_NEW_VOTERS))

```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**A4** The two regions with the most new voters are Central Maryland and the Capitol Region. 

```{r}
md_voters_2024 |>
  
mutate(REGION = case_when(
    COUNTY == "Garrett" ~ "Western Maryland",
    COUNTY == "Allegany" ~ "Western Maryland",
    COUNTY == "Washington" ~ "Western Maryland",
    COUNTY == "Frederick" ~ "Capitol Region",
    COUNTY == "Montgomery" ~ "Capitol Region",
    COUNTY == "Prince George's" ~ "Capitol Region",
    COUNTY == "Anne Arundel"~ "Central Maryland",
    COUNTY == "Baltimore City"~ "Central Maryland",
    COUNTY == "Baltimore County"~ "Central Maryland",
    COUNTY == "Carroll"~ "Central Maryland",
    COUNTY == "Harford"~ "Central Maryland",
    COUNTY == "Howard"~ "Central Maryland",
    COUNTY == "Calvert"~ "Southern Maryland",
    COUNTY == "Charles"~ "Southern Maryland",
    COUNTY == "St. Mary's"~ "Southern Maryland",
    COUNTY == "Kent"~ "Eastern Shore",
    COUNTY == "Queen Anne's"~ "Eastern Shore",
    COUNTY == "Talbot"~ "Eastern Shore",
    COUNTY == "Caroline"~ "Eastern Shore",
    COUNTY == "Dorchester"~ "Eastern Shore",
    COUNTY == "Wicomico"~ "Eastern Shore",
    COUNTY == "Somerset"~ "Eastern Shore",
    COUNTY == "Worcester"~ "Eastern Shore",
    COUNTY == "Cecil"~ "Eastern Shore" ),
    
  NEW_REGISTRATION = case_when(
          STATE_REGISTRATION_DATE - COUNTY_REGISTRATION_DATE == 0 
          ~ "NEW_VOTER")
)|>

group_by(REGION)|>
  filter(NEW_REGISTRATION == "NEW_VOTER") |>
  summarise(
     TOTAL_NEW_VOTERS = n()) |>
  arrange(desc(TOTAL_NEW_VOTERS))
  
```

**Q5** What's an additional question you could ask of this data?

**A5** Out of the new voter registrations how many registered as Dem V Republican or how many new voters per capita are there per region? 
