---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```
The library function loads an R library such as the tidyverse, providing access to the verbs and functions that library contains. You have to load a library before you can use its functions.

Load the Maryland WinRed contributions dataset by running the following codeblock. 

```{r}
maryland_winred_contributions <- read_rds("pre_lab_01.rds")
```

```{r}
#print will take a string and return it in the terminal
print("Hello World")
```
```{r}
glimpse(maryland_winred_contributions)
```
glimpse of all data, meta data

```{r}
head(maryland_winred_contributions)
```
shows first 6 rows

```{r}
maryland_winred_contributions |>
  group_by(date) |>
  summarise(
    count_contribs = n()

  )|>

arrange(count_classes)
```

group_by : function that takes input (name of column you're grouping by). columns are attributes-- chooses which attribute to focus on

summarize : does the math, counting n()





|>  (and)
group_by : groups by given input 
summarize: new col
n() : counts number of rows/entries that exist per grouped attribute
sum(): adds data
n_distinct: only counts entry once, even if it appears multiple times
unique() : gets rid of all other rows
 
 
 mutate : function that creates a new column based on math with other columns, for every 
 
lubridate

County %in% c("Montgomery", "Prince George's", "Frederick")


TRUE 

if the county name is in the column-- then assign it to new name

