---
title: "lab_03"
author: "Derek Willis"
date: "2024-08-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data

```{r}
# Load MD slice of FEC House candidate contribution data
md_contribs <- read_rds("data/lab_03.rds")
```

## Fix header issues

You'll need to fix one issue with the column names using janitor. Write a code block to do that, making sure that you save the results to the same dataframe.

```{r}
cleaned_md_contribs <- md_contribs |>
  clean_names()|>
  
  glimpse()
```


## Answer questions

Q1: Which seven cities in Maryland had the largest *number* of contributions from donors? How many did each have? To answer this question, you'll have to clean the messy "contributor_city" field.

Note: To answer this question, you'll need to see every variation of the "contributor_city" for the top seven cities, which you should identify through some initial analysis. Then find variations of them and change those variations into the most common version.

A1: BETHESDA	1177			
CHEVY CHASE	704			
BALTIMORE	692			
ANNAPOLIS	492			
POTOMAC	490			
SILVER SPRING	408			
ROCKVILLE	359	

```{r}
cleaned_md_contribs |>

### Top 7: Bethesda, Chevy Chase, Baltimore, Annapolis, Silver Spring, Rockville, POTOMAC
### SILVER SPRINGS (3),BALRIMORE,, BLATIMORE, POTAMIC,ROCKWELL

mutate(contributor_city = case_when(
    contributor_city == "SILVER SPRINGS" ~ "SILVER SPRING",
    contributor_city == "BALRIMORE" ~ "BALTIMORE", 
    contributor_city == "BLATIMORE" ~ "BALTIMORE", 
    contributor_city == "POTOMIC" ~ "POTOMAC",
    contributor_city == "ANAPOLIS" ~ "ANNAPOLIS",
    contributor_city == "ROCKWELL" ~ "ROCKVILLE",
    TRUE ~ contributor_city))|>

 group_by(contributor_city) |>
summarize(
  city_count = n()
  )|>
  arrange(desc(city_count)) 

```

Q2: Find the 5-digit ZIP code in Maryland that had the highest total amount of contributions (by dollar value). You will need to use mutate to ensure that you have a zip code column with the first five characters from the existing zip code column.

Where is that zip code and what kind of area is it, demographically? Which committee is the largest recipient of money from that zip code? How many of the top five recipients are Maryland candidates?

A2: ZIP code 20817 is located in central Maryland and covers a slightly less than average land area compared to other ZIP codes in the United States. It also has a slightly higher than average population density.

The people living in ZIP code 20817 are primarily white. The number of middle aged adults is extremely large while the number of seniors is extremely large. There are also a large number of families and a small number of single adults. The percentage of children under 18 living in the 20817 ZIP code is slightly higher than average compared to other areas of the country. Median Household Income	$170,293  -- https://www.unitedstateszipcodes.org/20817/

 (top_zip_code = case_when(
         contributor_zip == "20817" ~ "top_zip_contributor")
       )

```{r}
cleaned_md_contribs |>
  
mutate(contributor_zip = str_sub(contributor_zip, start=1L, end=5L),
       ) |>
  
  filter(contributor_zip == "20817")|>
group_by(committee_name) |>
summarize(
 contribution_amount = sum(contribution_receipt_amount)
)|>
arrange(desc(contribution_amount))



```

Q3: Let's look at top individual donors. Write code to generate a list of top donors using name and city, adding up the total amount of contributions. What does the original data say about the top two donors - who are they and what do they do? Do you see anything in the list of top 10 donors that would make you uncertain about the accuracy of these results?

A3: It says Melanie and Mark Jacobsen are the two top donors. According to her linkedin Melanie is a former SEC senior special counsel in the Office of the Chief Accountant. According to dccouplestherapist.com (ha) Mark is a licensed clinical social worker in Washington, DC, Maryland, and Virginia. 

Some of the top donors have variations of the same name, which might mean they are the same person, which means they should be combined, which would change the order of the top 10.

```{r}

cleaned_md_contribs |>
  group_by(contributor_name, contributor_city)|>
  
summarise(
 donation_amount = sum(contribution_receipt_amount)
)|>
  
  arrange(desc(donation_amount))

```
