---
title: "lab_06"
author: "derek willis"
date: "2024-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse and any other libraries you'll need
library(tidyverse)
library(lubridate)
library(janitor)

```

## Get Our PDF

We'll be working with the [Maryland monthly voter registration update](https://elections.maryland.gov/pdf/vrar/2024_09.pdf) from the State Board of Elections. You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to write a couple of paragraphs that summarize the changes in voter registration in Maryland during the month of September.

## Setup Tabula

Start Tabula, then go to <http://127.0.0.1:8080/> in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds.

This PDF has multiple possible tables to extract. We're going to make four dataframes: 

* new registrations by party -- check
* removals by party -- check
* changes in registration -- check changes_in_registration
* current registrations by county. --check  total active registrations by county

You will need to draw boxes around four sections and export four CSV files that you will load into R. In Tabula, draw a box around the borders of those tables and click the "Preview & Export Extracted Data" button for each one. The export should look pretty clean, but might be missing some column headers. You'll need to include total columns.

Export each CSV (each one should be called `tabula-2024-09.csv` by default, so you should rename them for clarity) to your lab_06/data folder.

From there, you will need to read in and clean up the data, and in particular the headers. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. The pre_lab has an example of how to do this!

## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers must not contain spaces and they should have meaningful names, not "x1" or something similar. The current registrations dataframe *MUST* include the county name, but should not include the columns under `Changes` and `Party Affiliation From`. How you do that is up to you (and could involve something outside R), but you can use `select()` with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns.

```{r}
# Party registrations
party_registration_2024 <- read_csv("data/new_registration_by_party_2024.csv")


party_registration_2024 <- party_registration_2024 |>
 rename('method'= ...1)|>
adorn_totals()
  
party_registration_2024
```

```{r}
# Removals
removals_2024 <- read_csv("data/removals_by_party_2024.csv")

removals_2024 <- removals_2024 |>
 filter(!is.na(...1))|>
 rename('method'= ...1)|>
adorn_totals()

removals_2024

```

```{r}
# Changes
changes_2024 <- read_csv("data/changes_in_registration_2024.csv")

changes_2024 <- changes_2024|> 
   rename('county'= ...1)|>
  clean_names()|>
  select(county, address, name)|>
  adorn_totals()


changes_2024

```

```{r}
# Current registrations

current_registrations_2024 <- read_csv("data/total_active_registration_by_county_2024.csv")

current_registrations_2024 <- current_registrations_2024 |>
  clean_names()|>
  select(county,dem,rep,nlm,lib,grn,unaf,oth,total)|>
adorn_totals()
  

current_registrations_2024
```

## Rinse, Repeat: September 2022 data

Do all of that again, for the [September 2022 voter registration update](https://elections.maryland.gov/pdf/vrar/2022_09.pdf), so we can compare the two. If you've done well, this is easy: just do the same steps in Tabula, export the files and use the same code to read in and clean up the data, *just remember to give these dataframes different names* so that you can compare September 2024 to September 2022.

To answer the following questions, you'll need to calculate percentages and do other calculations. Whether you do that by combining or joining dataframes or you write code to run separately on both is up to you.

```{r}
# Party registrations
party_registration_2022 <- read_csv("data/new_registration_by_party_2022.csv")


party_registration_2022 <- party_registration_2022 |>
 rename('method'= ...1)|>
  adorn_totals()
  
party_registration_2022
```

```{r}
# Removals
removals_2022 <- read_csv("data/removals_by_party_2022.csv")

removals_2022 <- removals_2022 |>
 filter(!is.na(...1))|>
 rename('method'= ...1)|>
adorn_totals()

removals_2024
```

```{r}
# Changes
changes_2022 <- read_csv("data/changes_in_registration_2022.csv")

changes_2022 <- changes_2022|> 
   rename('county'= ...1)|>
  clean_names()|>
  select(county, address, name)


changes_2022
```

```{r}
# Current registrations

current_registrations_2022 <- read_csv("data/total_active_registration_by_county_2022.csv")

current_registrations_2022 <- current_registrations_2022 |>
  rename('county'= ...1)|>
  clean_names()|>
  select(county,dem,rep,grn,unaf,lib,wcp,oth,total)|>
  adorn_totals()

  

current_registrations_2022
```

## Answer questions

Q1. Calculate the percentage of registered Democrats, Republicans and unaffiliated voters for the entire state in September 2024 and September 2022, and then calculate the percentage change for those three party categories (the formula for percentage change is easy to remember: (New-Old)/Old). How would you describe the changes in these partisan categories in a story? What's the most interesting thing about these changes that would be the lede of that story?

2022
total percent dem 54.08836	
total percent rep 24.00269	
total percent unaf 20.22216

2024
total percent dem 52.45109
total percent rep 23.85315	
total percent unaf 21.835

percent change 
dem  -1.401357%
rep  1.042942%
unaf 9.785723%



The most interesting: The amount of Maryland voters who registered as "unaffiliated" grew by almost 10% between 2022 and 2024.

Within the same time frame, the amount of Maryland voters registered as democrats decreased by 1.4% while Republican affiliations rose by 1%.


A1.

```{r}
#2022
current_registrations_2022 |>
  filter(county=='Total')|>
  group_by(county,dem,rep,unaf,total)|>
  summarise(
    percent_dem = (dem/total)*(100),
    percent_rep = (rep/total)*(100),
    percent_unaf = (unaf/total)*(100)
  )
current_registrations_2022
```

```{r}
#2024
current_registrations_2024 |>
  filter(county=='Total')|>
  group_by(county,dem,rep,unaf,total)|>
  summarise(
    percent_dem = (dem/total)*(100),
    percent_rep = (rep/total)*(100),
    percent_unaf = (unaf/total)*(100)
  )
current_registrations_2024
```

```{r}
#percent change
renamed_current_registrations_2024 <- current_registrations_2024 |>
  select(county, dem, rep, unaf)|>
  rename(dem_2024=dem, rep_2024=rep, unaf_2024=unaf)

registrations_2022_2024 <- current_registrations_2022 |>
  select(county, dem, rep, unaf)|>
  rename(dem_2022=dem, rep_2022=rep, unaf_2022=unaf)|>
  left_join(renamed_current_registrations_2024 , join_by(county)) 

registrations_2022_2024 |>
  filter(county=='Total')|>
 mutate(
   percent_change_dem = ((dem_2024-dem_2022)/(dem_2022))*100,
   percent_change_rep =((rep_2024-rep_2022)/(rep_2022))*100,
   percent_change_unaf =((unaf_2024-unaf_2022)/(unaf_2022))*100
 )|>

glimpse()

```


Q2. Among the Maryland jurisdictions with the most registered voters, which one had the biggest percentage increase in unaffiliated voters between 2022 and 2024? How did that county vote for governor in 2022 and for president in 2020? Contrast the partisan percentages of that county with the county that had the eighth-largest percentage increase in unaffiliated voters. You should display information from the data that shows the counties with the largest percentage increase first.

A2.

PG County had the largest percent increase in unaf voters between 2022 and 2024 at 17.2%.
For Governor, 89.23% ofPG county voted for Dem Wes Moore while for president, 85.90% of PG county voted for Dem Kamala Harris.
Frederick had the eighth-largest percent increase at 10.543469%.
For Governor, 53.46% of Frederick county voted for Dem Wes Moore while for president, 52.79% of Frederick county voted for Dem Kamala Harris.

```{r}

registrations_2022_2024|>
mutate(
   pc_unaf = ((unaf_2024-unaf_2022)/(unaf_2022))*100
)|>
  
 select(county,unaf_2022,unaf_2024,pc_unaf)|>
 arrange(desc(pc_unaf))

```

Q3. Do the party percentages of new registrants (% Dem, % Rep, % Unaffiliated) from September 2024 and September 2022 match the overall percentage of partisan registration in the state at those times?

In other words, did new registrants mirror the overall partisan composition of the electorate, or were they different? Do any of the methods of registration vary significantly from the overall partisan pattern? Which method is worthy of further exploration?

A3.
OVERALL PARTISAN REGISTRAION DATA
2022
total percent dem 54.08836	
total percent rep 24.00269	
total percent unaf 20.22216

2024
total percent dem 52.45109
total percent rep 23.85315	
total percent unaf 21.835

NEW REGISTRANTS
2022
percent new dem   51.63863
percent new rep   15.04665
percent new unaf  30.60096

2024
percent new dem  49.85496
percent new rep 18.23337 
percent new unaf 28.68156


OVERALL PARTISAN REGISTRAION percent change 
dem  -1.401357%
rep  1.042942%
unaf 9.785723%

NEW REGISTRANT percent change 
new dem -1.806671
new rep 23.24678
new unaf -4.672978

New registrations by party do not mirror overall composition. Especially in terms of the percent change in new republican voters. 


Based on the percent change between 2022 and 2024, Absentee voting has varied greatly compared to other methods.
The percent change of new voters requesting absentee ballots by party at registration is as follows:
dem  577.551020
rep  493.333333 
unaf    362.500000



```{r}
party_registration_2022 <- party_registration_2022 |>
  clean_names()|>
select(method, dem, rep,unaf,total)
  
party_registration_2024 <- party_registration_2024 |>
  clean_names()|>
select(method, dem, rep,unaf,total)
```

```{r}
#party registration 2022 percent
party_registration_2022|>
  mutate(
    percent_new_dem = (dem/total)*100,
    percent_new_rep = (rep/total)*100,
    percent_new_unaf = (unaf/total)*100
  )|>
filter(method=='Total')

#party registration 2024 percent
party_registration_2024|>
  mutate(
    percent_new_dem = (dem/total)*100,
    percent_new_rep = (rep/total)*100,
    percent_new_unaf = (unaf/total)*100
  )|>
filter(method=='Total')

```
```{r}
#party registration 2022 2024 percent change
renamed_party_registration_2024 <- party_registration_2024 |>
  select(method, dem, rep, unaf)|>
  rename(dem_2024=dem, rep_2024=rep, unaf_2024=unaf)

new_registrations_2022_2024 <- party_registration_2022 |>
  select(method, dem, rep, unaf)|>
  rename(dem_2022=dem, rep_2022=rep, unaf_2022=unaf)|>
  left_join(renamed_party_registration_2024 , join_by(method)) 

new_registrations_2022_2024 |>
  filter(method=='Total')|>
 mutate(
   percent_change_dem = ((dem_2024-dem_2022)/(dem_2022))*100,
   percent_change_rep =((rep_2024-rep_2022)/(rep_2022))*100,
   percent_change_unaf =((unaf_2024-unaf_2022)/(unaf_2022))*100
 )|>

glimpse()


```
```{r}
#Methods
new_registrations_2022_2024 |>
 mutate(
   percent_change_dem = ((dem_2024-dem_2022)/(dem_2022))*100,
   percent_change_rep =((rep_2024-rep_2022)/(rep_2022))*100,
   percent_change_unaf =((unaf_2024-unaf_2022)/(unaf_2022))*100
 )

  new_registrations_2022_2024 

```


Q4. Choose your own adventure! Using both the 2024 and 2022 data for one of the dataframes (changes, removals or new registrations), ask and answer a question that makes at least one comparison and requires code to answer. That could involve calculating a percentage, percentage change or the raw difference between two columns.

A4.Which party affiliation had the greatest percent of removals per year for 2022 and 2024?

2022: 57.78056


2024: 57.98707

```{r}

clean_removals_2022 <- removals_2022 |>
  clean_names()|>
  select(method,dem,rep,unaf,total)|>
rename(dem_2022=dem, rep_2022=rep, unaf_2022=unaf, total_2022=total)

clean_removals_2024 <- removals_2024 |>
  clean_names()|>
  select(method,dem,rep,unaf,total)|>
rename(dem_2024=dem, rep_2024=rep, unaf_2024=unaf, total_2024=total)


removals <- clean_removals_2022 |>
 left_join(clean_removals_2024,join_by(method))|>
filter(method=='Total')|>
mutate(
  dem_removal_2022= (dem_2022/total_2022)*100,
  rep_removal_2022= (rep_2022/total_2022)*100,
  unaf_removal_2022=( unaf_2022/total_2022)*100,
  dem_removal_2024= (dem_2024/total_2024)*100,
  rep_removal_2024= (rep_2024/total_2024)*100,
  unaf_removal_2024= (unaf_2024/total_2024)*100
)
  
removals


```

Q5. What's the best story idea or question you've seen as a result of the work you've done in this lab?

A5. The best question would be why has there been an increase in UNAF voters in Maryland and how could it affect election results? Is there a specific topic that has popped up in this time frame that has disillusioned people with their original party/ dividing people (ahem Israel v Palestine?)? How might this trend continue? How will it affect future election results and/or how each party will realign or adjust to re-attract voters?
