---
title: "lab_05"
author: "derek willis"
date: "2024-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# load the tidyverse
library(tidyverse)
```

## Getting Started

This week we'll be combining and joining data to make it more useful and to ask some questions about it. This week, we'll revisit our new Maryland voters data and try to join it with contributions to the Harris campaign from Maryland donors made via ActBlue. Can we find new voters who are not registered Democrats who gave to Harris after she became a presidential candidate on July 21? What, if anything, can we say about those people?

We'll be doing something similar to what Ence Morse did for [this Washington Post story](https://www.washingtonpost.com/politics/2024/09/24/kamala-harris-younger-donors/?pwapi_token=eyJ0eXAiOiJKV1QiLCJhbGciOiJIUzI1NiJ9.eyJyZWFzb24iOiJnaWZ0IiwibmJmIjoxNzI3MzIzMjAwLCJpc3MiOiJzdWJzY3JpcHRpb25zIiwiZXhwIjoxNzI4NzA1NTk5LCJpYXQiOjE3MjczMjMyMDAsImp0aSI6ImIwYWMxNTk4LWMwMDgtNDE5Ny04MjAyLWVmMWNjZWFmYWZiMiIsInVybCI6Imh0dHBzOi8vd3d3Lndhc2hpbmd0b25wb3N0LmNvbS9wb2xpdGljcy8yMDI0LzA5LzI0L2thbWFsYS1oYXJyaXMteW91bmdlci1kb25vcnMvIn0.3-s0tBkMV7Y9DD21CBsKMlaliGlOKT2egCO-UFT735Y), just on a smaller scale.

First, load two CSV contribution files in the `data` folder, one each for donations made in July and August, and then combine them into a single dataframe.

```{r}
# load the Harris contribution data files and combine them
harris_august <- read_csv("data/august_harris.csv")
harris_july <- read_csv("data/july_harris.csv")

harris_combined <- bind_rows(list(harris_august,harris_july))
View(harris_combined)

```

Next, let's read in the new voters (people who registered in 2024):

```{r}
# load the new voters data

  
new_voters <- read.csv("data/new_voters.csv")


```

Next join the Harris donations to the voter data. Remember, people's names are common, __so you'll want to use name columns and address information__. You'll need to figure out which columns to use from both dataframes. The new_voters data has both a single Address column and various parts of the address, such as HouseNumber. Make a plan for picking your join columns - try out some combinations - and save the results to a new dataframe.


Names: 
LastName FirstName MiddleName
last_name first_name middle_name
ResidentialZip
zip
ResidentialCity
city

Address

```{r}
# join the combined Harris file to the new voters
# you will need to determine what columns to use

  
 harris_and_new_voters <- harris_combined |> 

  inner_join(new_voters, join_by(
address_one==Address,
last_name==LastName,
first_name==FirstName,
city==ResidentialCity),
)



```

## Answer questions

Q1. Explain why you used the columns you did to make the join, and why that result is better than other ways you tried. Are there any problems with the join you made, such as the possibility of false positives?

A1.  I used address because each household has a unique address. Within a household people have different names, and there may be people with the same name and same address in different cities.

Q2. Based on your join, 
how many donations to Harris have come from people who registered this year? How many unique people does that represent, and which donors have given the largest _number_ of donations since July 20? 

A2. Harris received 326	donations from new voters. SOPHIA	HOFFER-PERKINS,ERIC	SNIPE and BRITTANY	MUSOLINO	donated 7 times each

```{r}
harris_and_new_voters |>
  
  group_by(VTR_ID, first_name, last_name)|>
summarise(
  donations = n()
)|>
  arrange(desc(donations))
```


Q3. What are the party and gender breakdowns of those donors? How many donations has Harris gotten from non-Democrats?

DEM	Female	160		
DEM	Male	123		
DEM	NA	11		
LIB	Male	2		
NLM	Male	1		
OTH	Female	2		
REP	Female	3		
REP	Male	1		
UNA	Female	8		
UNA	Male	14	

Harris received  31 donations from non-democrats


A3. 

```{r}

harris_and_new_voters |>
  group_by(Party, Gender)|>
summarise(
 donations = n()
)

```

Q4. Choose your own adventure: I want you to come up with a question you could ask of one of the dataframes we've created for this lab. It could involve just the new voters, or the donors or the joined dataframe, but it cannot be a variation of a previous question (so, for example, you can't ask, "How many Republicans have given to Harris"). First write the question and then try to answer it by writing code. If you're not sure of the exact code, try googling (or ask me for advice) but, at a minimum, describe what you think you'd have to do to answer it and what the results would look like. That can help.

A4:
How much did other parties donate, and which party donated the most?

UNA donated the most of any non DEM party with	$1294.50	.

```{r}


harris_and_new_voters |>
  filter(Party!="DEM")|>
group_by(Party)|>
summarise(
 donation_amount = sum(amount)
)|>

arrange(desc(donation_amount))

```



