---
title: "lab_04"
author: "Derek Willis"
date: "2024-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(janitor)
library(lubridate)

```

For this lab, we want to investigate spending by the two leading Senate candidates in Maryland during this election, Angela Alsobrooks and Larry Hogan. Our goal is to try and standardize the purpose of the expenditures so that we can compare the two campaigns. We have the data, but it's messy - purpose descriptions are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of campaign expenditures located in the data folder called "md_senate_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
md_senate_expenses <- read_csv("data/md_senate_expenses.csv")

```

## Answer questions

**Q1.** You've been assigned to compare the spending priorities of the two campaigns, but they describe their spending differently using the `purpose` column. Using OpenRefine, create a project using the original CSV file, make a copy of the `purpose` column called `purpose_clean` and then standardize it, focusing on the purposes that appear most often or have the largest total amounts. You can transform the data as you want - making all records the same case, for example, to help. The goal is to be able to group together as many similar purpose values, so you can choose to call all transportation-related expenses the same thing.

How do you know when you're done? It's a judgment call, but you should be close to 100 unique values for `purpose_clean`, and maybe even below that number.

Then, when you are done standardizing `purpose_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.

Then, using that `purpose_clean` column, write code to find the largest amounts of spending on categories by both campaigns.

**A1.** Alsobrooks 
Advertising	4122402.70			
Salary	822014.96			
Consulting 699441.46	

Hogan
Advertising	2178145.48			
STAFFING	481109.49			
Fundraising	315492.06	





```{r}
md_senate_expenses_clean <- read_csv("data/md-senate-expenses-clean.csv")

  md_senate_expenses_clean |>
 clean_names()|>
  filter(candidate == "Alsobrooks") |>
group_by(purpose_clean)|>
  
summarise(
  expense = sum(amount)
)|>
  
  
  arrange(desc(expense))

```
```{r}

  md_senate_expenses_clean |>
 clean_names()|>
  filter(candidate == "Alsobrooks") |>
group_by(purpose_clean)|>
  
summarise(
  expense = sum(amount)
)|>
  
  
  arrange(desc(expense))

```
```{r}

  md_senate_expenses_clean |>
 clean_names()|>
  filter(candidate == "Hogan") |>
group_by(purpose_clean)|>
  
summarise(
  expense = sum(amount)
)|>
  
  
  arrange(desc(expense))

```


**Q2.** Let's do some research on the top purposes from A1. Feel free to search the Web to understand those terms better if need be, and then write a description comparing and contrasting the two campaigns' largest categories of spending.

**A2.** Both candidates spent the most money on advertising, which makes sense given that these are campaigns. They would like to bring as much attention to their candidates as possible. 
Second is staffing and salaries, which is probably something that could have been grouped together retrospectively. The two campaigns probably just called the two things something different. Then, I have Consulting third for Alsobrooks and fundraising third for Hogan. I'm not sure about this, but I imagine fundraising costs are probably something like event costs or the cost of reaching out to donors other than advertising. With Alsobrooks and consulting, I imagine most of the consultation is on what kind of advertising they should focus on and how they should do it. Therefore, despite some slight different names, I think the campaign's largest categories of spending are pretty similar/ However, Alsobrooks spent double in each top category I mentioned.


**Q3.** Pick one of the top categories for each campaign and write code that shows how that spending has changed over time, if at all. You'll need to use lubridate for this. Write a couple of sentences describing that pattern.

**A3.** For the Hogan campaign, I selected the STAFFING category. There are only 5 months worth of data, but there is a very big change in the amount of spending in this category. In February 2024, only $19949.01	was spent on Staffing while in June of the same year, $148589.29 was spent. For the Alsobrooks campaign I focused on Salary, which I think this the equivalent to Hogan's staffing category. For this, there are even fewer months to compare, but the change is not as drastic. The Alsobrooks campaign started with high salary expenses $77651.81	, but stayed fairly stable, while the Hogan campaign started much smaller than this then ended up paying more than double what her campaign did in the last month.



```{r}
 md_senate_expenses_clean |>
 clean_names()|>
   filter(candidate == "Hogan", purpose_clean == "STAFFING") |>
  mutate(month = floor_date(date, "month")
            )|>
group_by(purpose_clean, month)|>
  
summarise(
  expense = sum(amount)
)





```
```{r}
 md_senate_expenses_clean |>
 clean_names()|>
   filter(candidate == "Alsobrooks", purpose_clean == "Salary") |>
  mutate(month = floor_date(date, "month")
            )|>
group_by(purpose_clean, month)|>
  
summarise(
  expense = sum(amount)
)



```


**Q4.** Which committee has the larger payroll or salary spending so far? How many unique recipients of payroll/staff spending are there for each candidate?

**A4.**
Alsobrooks has had larger salary spending by almost double. Hogan has paid $481,109.5	while Alsobrooks has paid	$822,015. Alsobrooks has 23 unique payroll recipients under Salary while Hogan has one under STAFFING. Each of Hogan's staffing charges goes to Purple Surfboard, which according to the Washington Post is a political consulting firm. So either they source their staffers from that firm or I chose the wrong category of spending, but I cannot find any other categories that seem to target specific recipient other than reimbursement charges, so I'm not changing it.


```{r}
md_senate_expenses_clean |>
 clean_names()|>
   filter(purpose_clean == "Salary" | purpose_clean == "STAFFING") |>
  group_by(candidate, purpose_clean)|>
  
summarise(
  expense = sum(amount)
)


```
```{r}
md_senate_expenses_clean |>
 clean_names()|>
   filter( purpose_clean == "Salary" | purpose_clean == "STAFFING") |>
  
  group_by(purpose_clean)|>
summarise(
  payroll = n_distinct(last_name)
)

```

**Q5** What's the most interesting story idea you've found in this data?

**A5**
  I think the most interesting story has been the stark contrast between staff payroll/salary between the two campaigns. It is interesting to see how they differ in terms of overall expense, expense per month and how many people are on the payroll. It could be interesting to see who on the payroll gets paid the most v. the least. It could also be interesting to get more perspective into campaign salaries for these types of races in general to see how they compare ot other races or how these Maryland races have changed over time. It is also important to look at how successfull the campaigns were in contrast to how much they paid their staff. 
  Also, if I am in fact correct about all of the staffing funds going toward Purple Surfboard, it would be interesting to try to get further information about why and figure out if from there we can break down how much each staffer received from the campaign, because each staffing charge going to the firm, makes it impossible to know how many people were actually on the payroll. Why is it like this, what are we missing because of it and why do they want it that way?
